---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Proximum

<!-- badges: start -->
[![R-CMD-check](https://github.com/agostinognasso/Proximum/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/agostinognasso/Proximum/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/agostinognasso/Proximum/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/agostinognasso/Proximum/actions/workflows/test-coverage.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The proximity matrix of a tree ensemble records how often two observations fall
in the same leaf. `randomForest` will compute it for you and then leave you to
it: there is no way in R to test whether two forests represent the data the same
way, to ask how many trees a stable proximity needs, or to compute the thing at
all when `n` is large.

`Proximum` treats the proximity matrix as a first-class statistical object
rather than as a by-product of the model: one extractor across engines, in-bag
and out-of-bag definitions, the transformations and metric diagnostics that go
with them, and permutation inference for comparing two matrices or partitioning
one. Stability diagnostics, scalable approximations and the plots are planned;
the functions that will provide them are documented and raise an error naming
their release.

It is the methodological layer underneath
[`e2tree`](https://cran.r-project.org/package=e2tree), which reconstructs a
forest's behaviour from exactly this similarity structure.

## Installation

Not on CRAN yet. Install from GitHub:

```r
# install.packages("pak")
pak::pak("agostinognasso/Proximum")
```

## Usage

```{r usage, eval = requireNamespace("randomForest", quietly = TRUE)}
library(Proximum)

set.seed(1)
rf <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 200, keep.inbag = TRUE
)

px <- proximity(rf, newdata = iris)
summary(px)
```

Switching to `type = "oob"` removes the optimistic bias of the in-bag
definition, at the cost of positive semi-definiteness: the out-of-bag
proximity is not a kernel, and `summary()` says so. See
`vignette("Proximum-intro")`.

## Status

Early development. Extraction from `randomForest` and `ranger`, the
dissimilarity transforms, the Euclidean diagnostics, the positive semi-definite
repair and the whole inference layer all work. The stability and scalability
layers and the plots are declared, documented and not yet implemented, and
calling one of those functions raises an error that says which phase it belongs
to.

Every number quoted in the documentation comes from a script in
`inst/simulations/`, and those are meant to be re-run rather than believed.

| Phase | Content | State |
|---|---|---|
| F1 | `randomForest` and `ranger`, `proximity` object, transforms, `make_psd()` | mostly done |
| F2 | `mantel_test()`, `cka()`, `rv_coefficient()`, `permanova()`, `protest()` | done |
| F3 | Sparsity, Nystrom, stability, `n_trees_required()` | declared |
| F4 | `autoplot()`, vignettes, pkgdown | declared |
| F5 | CRAN, JSS paper | not started |

The proximity is computed as a sparse Gram matrix, `P = Z Z'/B` with `Z` the
leaf indicator, rather than by looping over trees. On 300 trees and 1500
observations that is 0.16 s instead of 13 s, and the same identity is why the
in-bag proximity is positive semi-definite while the out-of-bag one, an
elementwise quotient of two Gram matrices, is not.

## Related work

* [`e2tree`](https://cran.r-project.org/package=e2tree): explains a forest with
  a single tree, built on the same similarity structure.
* [`rankimp`](https://github.com/agostinognasso/rankimp): which variables drive
  the representation `Proximum` describes.
* [`vegan`](https://cran.r-project.org/package=vegan): the reference
  implementation of Mantel, PERMANOVA and Procrustes on ordinary
  dissimilarities. It cannot consume a matrix with undefined pairs, which is
  why this package implements them natively and uses `vegan` only in its
  tests, to check that the two agree where both apply.
