---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

# Proximum

<!-- badges: start -->
[![R-CMD-check](https://github.com/agostinognasso/Proximum/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/agostinognasso/Proximum/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/agostinognasso/Proximum/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/agostinognasso/Proximum/actions/workflows/test-coverage.yaml)
[![coverage](https://img.shields.io/badge/coverage-%E2%89%A595%25-brightgreen)](https://github.com/agostinognasso/Proximum/actions/workflows/test-coverage.yaml)
[![R package version](https://img.shields.io/github/r-package/v/agostinognasso/Proximum?label=version)](https://github.com/agostinognasso/Proximum/blob/main/DESCRIPTION)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Statistical analysis of the proximity matrices that tree ensembles produce.

## The problem

Fit a random forest and you get predictions, and you get an importance table.
You also get, whether you ask for it or not, an answer to a different question:
which observations does this model consider to be the same kind of thing? Two
observations that keep landing in the same leaf are, as far as the forest is
concerned, alike. Collect that over every tree and you have the **proximity
matrix**

$$P_{ij} = \frac{1}{B}\sum_{b=1}^{B} \mathbb{I}\big[\ell_b(x_i) = \ell_b(x_j)\big],$$

with $\ell_b(x)$ the leaf of tree $b$ that $x$ reaches. It is a full
description of the geometry the model learned, and it is the only part of a
forest that talks about the observations rather than about the variables.

`randomForest` will compute it if you pass `proximity = TRUE`, hand you the
matrix, and stop. There is no way in R to ask of it any of the questions that
make it a statistical object rather than a picture:

- Do two forests represent the data the same way? Two engines, two random
  seeds, this year's model against last year's.
- How much of the structure is explained by the response, and how much by
  something the model never saw?
- Is it stable? A matrix that changes when you refit is not evidence of
  anything, and nobody says how many trees it takes.
- What do you do when $n$ is large? The matrix is $n \times n$: 31 MB at
  $n = 2{,}000$ and about 760 MB at $n = 10{,}000$.
- Is it even a valid kernel? The out-of-bag definition, which is the one that
  removes the optimistic bias, is **not** positive semi-definite, and every
  method downstream that assumes a kernel is quietly wrong on it.

## What the package does instead

One extractor across engines, both definitions, the transforms and metric
diagnostics that go with them, permutation inference for comparing matrices and
partitioning one, two storage forms for when the matrix does not fit, a
stability layer, and five plots. Every number in the documentation comes from a
simulation study shipped in `inst/simulations/`, and several of those studies
exist because they contradicted something that had been written down first.

## Installation

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("agostinognasso/Proximum")
```

## A worked example, end to end

The package ships a synthetic consumer loan portfolio. It is generated rather
than collected, and `?loans` says exactly what was built into it and why; the
short version is that public credit panels with a protected attribute left in
them do not exist, and one of the questions below needs one.

```{r data}
library(Proximum)
library(randomForest)

str(loans, give.attr = FALSE, vec.len = 2)
```

### The proximity, and what it is

`as_proximity()` takes a fitted ensemble and the data it was fitted on.

```{r proximity}
predictors <- setdiff(names(loans), c("applicant_group", "vintage"))

set.seed(1)
book <- loans[sample(nrow(loans), 400), ]
rf <- randomForest(default ~ ., data = book[, predictors], ntree = 500)

px <- as_proximity(rf, newdata = book[, predictors])
px
```

The name is `as_proximity()` and not `proximity()` because `e2tree`, the
package most likely to be attached alongside this one, exports a `proximity()`
of its own. The `as_` prefix also says what it does: it coerces a fitted
ensemble into an object of class `proximity`, the way `as.dist()` coerces into
a `dist`.

`summary()` reports what the matrix looks like and, more usefully, whether the
dissimilarity it induces can be embedded in a Euclidean space at all.

```{r summary}
summary(px)
```

### The one thing everybody gets wrong

The in-bag proximity above averages over *all* trees, including the trees that
were fitted on $i$ and $j$. Those trees have seen both observations and are
inclined to separate them correctly, which inflates the proximity of same-class
pairs. Restricting the average to the trees where both were out-of-bag removes
that bias, and costs the geometry:

```{r geometry}
rf_oob <- randomForest(default ~ ., data = book[, predictors], ntree = 500,
                       keep.inbag = TRUE)
px_oob <- as_proximity(rf_oob, newdata = book[, predictors], type = "oob")

c(inbag = summary(px)$euclidean, oob = summary(px_oob)$euclidean)
```

This is not an accident of the data. Stack the leaf indicators into $Z$: the
in-bag proximity is $P = ZZ^{\top}/B$, a Gram matrix, hence positive
semi-definite. The out-of-bag one divides each entry by the number of trees in
which that *pair* was jointly out of bag, making it a Hadamard quotient of two
Gram matrices, which need not be positive semi-definite and here is not.
Debiasing the estimate costs the geometry. `make_psd()` will repair it and
record what it did, and `summary()` refuses to let the question pass silently
either way.

### Looking at it

Three views of the same object, and which one answers a question depends on the
question.

```{r heatmap, fig.width = 5.5, fig.height = 4.6}
autoplot(px, type = "heatmap")
```

The rows and columns are ordered by a seriation of the induced dissimilarity,
so the blocks the ensemble learned line up along the diagonal instead of being
scattered by the order the rows arrived in.

```{r mds, fig.width = 6.5, fig.height = 4.4}
autoplot(px, type = "mds", colour = book$default)
```

The colouring is yours to pass. A `proximity` object carries the engine, the
number of trees and the definition used, and nothing about the response, so a
plot that coloured by outcome on its own would be inventing the outcome.

```{r network, fig.width = 6.5, fig.height = 5}
set.seed(11)
autoplot(px, type = "network", threshold = 0.2)
```

The network keeps the pairs above a threshold and reads the communities of the
graph that remain. Those communities are the forest's own clustering, recovered
from the proximity rather than imposed from outside;
`vignette("credit-scoring-case")` asks whether they are risk segments and finds
one that defaults at close to nine in ten against a book rate of one in six.

### Comparing two forests

This is the question the package was written for. Fit a second forest, on the
same data, differing only in the seed, and ask whether the two represent the
borrowers the same way.

```{r mantel}
rf2 <- randomForest(default ~ ., data = book[, predictors], ntree = 500)
px2 <- as_proximity(rf2, newdata = book[, predictors])

mantel_test(px, px2, n_perm = 999)
```

Two refits of the same forest agree almost perfectly, which is the answer you
would hope for and the reference level everything else is read against. Change
what the forest is allowed to look at, and it moves:

```{r mantel-thin}
thin <- randomForest(default ~ score + utilisation, data = book, ntree = 500)
mantel_test(px, as_proximity(thin, newdata = book), n_perm = 999)$statistic
```

Three statistics are offered and they do not measure the same thing.
`mantel_test()` correlates the pairwise values and runs the whole range;
`cka()` aligns the centred kernels; `rv_coefficient()` is the matrix
correlation. Measured over 20 replications at $n = 300$, on forests with
*nothing* in common at all, Mantel reads 0.000 and CKA reads 0.670, and CKA
cannot separate replicates of one ensemble (0.991) from forests fitted to
different predictors with the same response (0.959). Use Mantel unless the
question really is about the kernels.

### Partitioning the variation

`permanova()` puts the proximity on the left of a design and asks how much of
it each term accounts for. The interesting use is a term the model never
received:

```{r permanova}
permanova(px, ~ default + applicant_group, data = book, n_perm = 999)
```

`applicant_group` is a protected attribute. It was not in the formula, and in
the generating process it affects the default probability nowhere at all. It
does shift the income and the score, which *are* in the formula, so the forest
separates the groups anyway, and this is how much: a small share, reliably not
zero, measured on a model that never saw the attribute. That is a fairness
diagnostic that does not require the model to have used the thing you are
worried about.

### How many trees does this need?

A proximity matrix that moves when you refit is not evidence of anything.
`n_trees_required()` grows the ensemble in disjoint blocks and stops when the
entries stop moving between them.

```{r trees}
required <- n_trees_required(rf, book[, predictors], eps = 0.15)
required
```

```{r trees-plot, fig.width = 6.5, fig.height = 4}
autoplot(required)
```

The criterion falls as a power of the block size, and the exponent is fitted
rather than assumed: over four generating processes and both engines it lay
between $-0.55$ and $-0.60$ with $R^2$ of at least 0.98, so it is a clean power
law whose exponent is not the $-0.5$ a plain standard error would give. The
answer is an ordinary integer and goes on behaving as one, so it can be handed
straight back to the engine that raised the question:

```{r trees-integer}
required + 100L
```

`stability()` asks the same question of replicates you already hold, and
assumes nothing about where they came from.

### When the matrix does not fit

Two storage forms, and they are not interchangeable.

```{r storage}
sparsify(px, threshold = 0.05)
nystrom(rf, book[, predictors], landmarks = 100)
```

`sparsify()` thresholds and stores the result sparsely. It saves memory and
**does not survive being used**: every statistic here runs on
$\sqrt{1 - P}$ or on the doubly centred matrix, and across every sample size
measured those had between 99.5 and 99.9 per cent, and exactly 100 per cent, of
their entries non-zero. So the inference layer refuses a sparse object by name
rather than densifying it behind you.

`nystrom()` is the one that survives. It stores an $n \times r$ factor whose
cross-product is the approximation, so the $n \times n$ matrix is never formed
going in or coming out: 40 MB instead of 760 at $n = 10{,}000$ with 500
landmarks. Its entries move much more than its geometry does, which is the
point, since the geometry is what it is for: at $n = 400$ with 200 landmarks,
20.7 per cent relative error on the entries and 4.9 per cent on the
configuration `embedding()` returns.

### Closing the loop with `e2tree`

`e2tree` fits a single explainable tree to the dissimilarity a forest induces.
It takes that dissimilarity as an argument, which is exactly what this package
produces, and returns a tree whose own partition comes back as a proximity:

```{r e2tree, eval = requireNamespace("e2tree", quietly = TRUE)}
explanation <- e2tree::e2tree(default ~ ., data = book[, predictors],
                              D = as_dissimilarity(px), ensemble = rf)
mantel_test(px, as_proximity(explanation), n_perm = 999)$statistic
```

That correlation is how much of the ensemble's geometry the single tree
reproduces, which is the question `e2tree` exists to answer and the one thing
it cannot ask of itself.

### What the example showed

The forest arranged 400 borrowers into blocks; one of those blocks is an
interaction no coefficient would have shown you. Two refits of the same forest
agreed almost perfectly and a forest given two predictors instead of ten did
not, so the statistic separates the cases it needs to separate. A protected
attribute that the model never received still accounted for a measurable share
of the arrangement. And the whole thing was checked for stability, for the
number of trees it rests on, and for whether it is a kernel at all.

## The theory underneath

**The Gram identity.** With $Z$ the stacked leaf indicator, $P = ZZ^{\top}/B$.
This settles three things at once: the in-bag proximity is positive
semi-definite, so $\sqrt{1-P}$ is Euclidean; the out-of-bag one is an
elementwise quotient of two such products and is not; and the matrix can be
computed as one sparse cross-product instead of a loop over trees. On 300 trees
and 1,500 observations that is 0.16 s against 13 s.

**Classical scaling and the kernel are the same computation.** With
$D^2 = \mathbf{1}\mathbf{1}^{\top} - P$ and $J$ the centring operator,
$J\mathbf{1} = 0$ kills the first term and
$G = -\tfrac12 J D^2 J = \tfrac12 J P J$. So `embedding()` never has to form
the dissimilarity, and the same identity is what lets it read a configuration
off a Nystrom factor at $O(nr^2)$ instead of $O(n^3)$.

**Permutation, not asymptotics.** Every entry of a proximity matrix shares an
observation with $2(n-2)$ others, so the entries are nowhere near independent
and no closed-form null applies. All the tests here permute the observations,
which permutes rows and columns together and preserves that dependence.

**Undefined is not zero.** An out-of-bag pair that was never jointly out of bag
is `NA`, not `0`. Storing a zero would assert that the two observations never
share a leaf, when what the forest reported is that it never had the chance to
look. Every statistic is computed on the pairs where all matrices involved are
defined, and reports how many those were.

## Every exported function

| | |
|---|---|
| `as_proximity()` | the extractor, with methods for `randomForest`, `ranger` and `e2tree` |
| `as_dissimilarity()`, `as.dist()` | $\sqrt{1-P}$ or $1-P$ |
| `is_euclidean()`, `double_centre()` | the metric diagnostics |
| `make_psd()` | clip, flip or shift an indefinite proximity onto the PSD cone |
| `embedding()` | the classical scaling configuration, direct or from a Nystrom factor |
| `mantel_test()`, `cka()`, `rv_coefficient()` | compare two proximity matrices |
| `permanova()` | partition one across the terms of a design |
| `protest()` | superimpose two configurations, Procrustes |
| `sparsify()`, `nystrom()` | the two storage forms |
| `stability()`, `n_trees_required()` | how much it moves, and how many trees it takes |
| `autoplot()` | five methods, one per class |
| `loans` | the synthetic portfolio the case study runs on |

## What has been measured

Four studies in `inst/simulations/`, seeded and re-runnable, and the
documentation quotes them rather than the other way round. Several exist
because a claim written down first turned out to be false when measured:

- `?sparsify` said most pairs never share a leaf and the thresholded matrix is
  typically very sparse. The exact zeros are 14.3 per cent of pairs at
  $n = 200$ and 58.5 per cent at $n = 1600$. Replaced by the table.
- `n_trees_required()` shipped with a default of `eps = 0.01`, which needs
  about 50,000 trees and was unreachable from any ensemble anyone would fit.
- The criterion was documented as falling like $1/\sqrt{B}$. The measured
  exponent is $-0.56$, and the projection now fits it instead of assuming it.
- Stratified landmarks were said to help. On the whole matrix they do not at
  all; on the rows they were designed for they do, and only when the budget is
  small relative to how rare the class is.
- The seriation behind the heatmap was documented as deterministic. It is not,
  on a proximity: the matrix is $k/B$ and takes at most $B+1$ distinct values,
  so the ordering breaks the ties at random and moves with the seed.

## What this does not protect you from

The proximity is a description of the model, not of the data. Everything here
is a statement about how one fitted ensemble arranges one sample; a different
forest on the same data arranges it differently, which is precisely what
`stability()` is for measuring rather than a caveat to be waved away.

A Mantel correlation is not a hypothesis about the world. It tests exchangeability
of the observation labels, and rejecting that null tells you the two matrices
share structure, not that the structure means what you hoped.

`permanova()` on a proximity is a diagnostic and not a fairness certificate. A
share of variation is one number about one representation.

And the out-of-bag proximity is not a kernel. Every method downstream that
assumes one, `cka()` included, refuses it rather than returning something
plausible, and `make_psd()` is a choice you make and not a repair that happens
to you.

## Documentation

- `vignette("Proximum-intro")`: proximity matrices as statistical objects, the
  in-bag and out-of-bag definitions and what separates them, the three views.
- `vignette("comparing-forests")`: the inference layer in full.
- `vignette("large-n")`: sparsification, Nystrom, stability, and how to choose.
- `vignette("credit-scoring-case")`: the three questions above, worked.

## Status

Early development, pre-CRAN. Phases F1 to F4 are done: extraction from
`randomForest`, `ranger` and `e2tree`, the transforms and diagnostics, the
positive semi-definite repair, the inference layer, the scalability and
stability layers, and the plots. What is declared and not provided is the
`streaming` argument that `vignette("large-n")` describes and says so about.

| Phase | Content | State |
|----|----|----|
| F1 | `as_proximity()`, the `proximity` object, transforms, `make_psd()` | done |
| F2 | `mantel_test()`, `cka()`, `rv_coefficient()`, `permanova()`, `protest()` | done |
| F3 | `sparsify()`, `nystrom()`, `embedding()`, `stability()`, `n_trees_required()` | done |
| F4 | `autoplot()`, the vignettes, the `loans` data | done |
| F5 | CRAN, JSS paper | not started |

## Related work

- [`e2tree`](https://cran.r-project.org/package=e2tree): explains a forest with
  a single tree built on the same similarity structure. It consumes what this
  package produces, and `as_proximity()` reads its result back.
- [`rankimp`](https://github.com/agostinognasso/rankimp): which variables drive
  the representation `Proximum` describes.
- [`vegan`](https://cran.r-project.org/package=vegan): the reference
  implementation of Mantel, PERMANOVA and Procrustes on ordinary
  dissimilarities. It cannot consume a matrix with undefined pairs, which is
  why this package implements them natively and uses `vegan` only in its tests,
  to check that the two agree where both apply.

## References

Breiman, L. (2001). Random forests. *Machine Learning*, 45(1), 5-32.

Anderson, M. J. (2001). A new method for non-parametric multivariate analysis
of variance. *Austral Ecology*, 26(1), 32-46.

McArdle, B. H. and Anderson, M. J. (2001). Fitting multivariate models to
community data. *Ecology*, 82(1), 290-297.

Cortes, C., Mohri, M. and Rostamizadeh, A. (2012). Algorithms for learning
kernels based on centered alignment. *JMLR*, 13, 795-828.

Williams, C. and Seeger, M. (2001). Using the Nystrom method to speed up kernel
machines. *NeurIPS 13*.

Bar-Joseph, Z., Gifford, D. K. and Jaakkola, T. S. (2001). Fast optimal leaf
ordering for hierarchical clustering. *Bioinformatics*, 17, S22-S29.

## License

MIT, (c) Agostino Gnasso.
