---
title: "Comparing forests through their proximities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing forests through their proximities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_rf <- requireNamespace("randomForest", quietly = TRUE)
```

Two forests fitted on the same data with different hyperparameters represent
that data differently. The question "how differently?" has no answer in the
usual toolkit, because the objects being compared are $n \times n$ matrices
rather than vectors of predictions.

## Two forests to compare

A shallow forest and a deep one, on the same rows. The shallow one is capped at
four leaves per tree, so it can only carve the data into coarse blocks; the
deep one is free to isolate individual observations.

```{r setup, eval = has_rf}
library(Proximum)
library(randomForest)

set.seed(1)
rows <- sample(nrow(iris), 60)
df <- iris[rows, ]

set.seed(2)
rf_shallow <- randomForest(Species ~ ., data = df, ntree = 300, maxnodes = 4,
                           keep.inbag = TRUE)
set.seed(3)
rf_deep <- randomForest(Species ~ ., data = df, ntree = 300, keep.inbag = TRUE)

px1 <- proximity(rf_shallow, newdata = df)
px2 <- proximity(rf_deep, newdata = df)
px1
```

## Correlation between two proximity matrices

The Mantel test correlates the off-diagonal entries of two matrices and gets its
null distribution by permuting the rows and columns of one of them together,
which preserves the dependence induced by the matrix structure as an ordinary
permutation of the entries would not.

That is not a technicality. Each entry of a proximity matrix shares an
observation with $2(n-2)$ others, so the $n(n-1)/2$ entries are a long way from
independent. Referring the correlation to the usual table would reject almost
whatever you fed it.

```{r mantel, eval = has_rf}
mantel_test(px1, px2, n_perm = 999)
```

The result carries the number of pairs it used. On an in-bag proximity that is
every pair; on an out-of-bag one it can be fewer, because a pair that was never
jointly out of bag has no proximity to correlate. The test uses what is defined
and reports how much that was.

```{r oob, eval = has_rf}
oob <- proximity(rf_deep, newdata = df, type = "oob")
mantel_test(px1, oob, n_perm = 999)$parameter
```

## Alignment of representations

`cka()` treats each proximity matrix as a kernel and measures the alignment of
the two implied feature spaces. It is invariant to isotropic rescaling and to
orthogonal transformation, which is why it has become the standard way of
comparing learned representations in the deep learning literature. Applying it
to forests puts ensemble explainability and representation similarity on the
same footing.

It is a coefficient and not a test: there is no p-value, and a high alignment
is not evidence of anything by itself. `mantel_test()` is where the evidence is.

```{r cka, eval = has_rf}
cka(px1, px2)
rv_coefficient(px1, px2)
```

The two differ only in the centring. `cka()` double-centres each matrix first,
which is what removes the mean similarity any two kernels on the same
observations share whether or not they have learned the same structure.

### Out-of-bag matrices have to be repaired first

An out-of-bag proximity is not positive semi-definite, and no number of trees
repairs it: its entries are ratios whose denominators count only the trees where
each pair was jointly out of bag, and those denominators differ across pairs.
An alignment computed on it is not an alignment between kernels, so `cka()`
refuses it.

The refusal is not about the number leaving `[0, 1]`. Over 600 out-of-bag
comparisons it never did. It is that the number is quietly too low: the
negative eigenvalues subtract from the numerator, so the uncorrected alignment
understated the corrected one in every one of those 600 comparisons, by 0.089
on average and by as much as 0.154.

```{r refuse, error = TRUE, eval = has_rf}
cka(oob, oob)
```

`make_psd()` is the repair, and it belongs to you rather than to `cka()`,
because which correction to use and what it costs is a decision about your data.

```{r repair, eval = has_rf}
cka(make_psd(px2), make_psd(oob))
```

## Still to come

`permanova()`, which will partition the variation in the induced dissimilarity
across the terms of a formula, and `protest()`, which will superimpose the two
multidimensional scaling configurations, are the rest of phase F2 and are not
implemented. So is the motivating case underneath all of this: `e2tree`
approximates a forest with a single tree, and whether that tree preserves the
forest's view of the data is exactly a question about two proximity matrices.
That comparison needs a `proximity()` method for `e2tree` objects, which does
not exist yet either.
