---
title: "Comparing forests through their proximities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing forests through their proximities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** The inference layer is scheduled for phase F2 of the roadmap. This
> vignette fixes the interface and the questions it is meant to answer; the code
> below does not run yet.

Two forests fitted on the same data with different hyperparameters represent
that data differently. The question "how differently?" has no answer in the
usual toolkit, because the objects being compared are $n \times n$ matrices
rather than vectors of predictions.

## Correlation between two proximity matrices

The Mantel test correlates the off-diagonal entries of two matrices and gets its
null distribution by permuting the rows and columns of one of them together —
which preserves the dependence induced by the matrix structure, as an ordinary
permutation of the entries would not.

```{r}
library(Proximum)

px1 <- proximity(rf_shallow, newdata = df)
px2 <- proximity(rf_deep, newdata = df)

mantel_test(px1, px2, n_perm = 999)
```

## How much of the structure is the response?

```{r}
permanova(px1, ~ y + region, data = df)
```

## Alignment of representations

`cka()` treats each proximity matrix as a kernel and measures the alignment of
the two implied feature spaces. It is invariant to isotropic rescaling and to
orthogonal transformation, which is why it has become the standard way of
comparing learned representations in the deep learning literature. Applying it
to forests puts ensemble explainability and representation similarity on the
same footing.

```{r}
cka(px1, px2)
protest(px1, px2)
```

## Comparing a forest with its explanation

The motivating case: `e2tree` approximates a forest with a single tree. Whether
that tree preserves the forest's view of the data is exactly a question about
two proximity matrices.

```{r}
px_forest <- proximity(rf, newdata = df)
px_e2tree <- proximity(e2tree_fit, newdata = df)

mantel_test(px_forest, px_e2tree)
```
