---
title: "A credit scoring case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A credit scoring case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** Placeholder. The case study lands with the inference layer
> (phase F2) and the scalability layer (phase F3), since a realistic credit
> portfolio needs both.

Credit scoring is the running example across `proxima`, `rankimp` and `e2tree`,
for three reasons. The regulatory context makes explanation a requirement rather
than a nicety; the data are large enough that the $O(n^2)$ cost of a proximity
matrix bites; and the predictors are correlated enough that different importance
measures genuinely disagree, which is what `rankimp` exists to handle.

## Questions this vignette will answer

1. Do the borrowers the forest treats as similar form recognisable risk segments,
   or does the proximity structure cut across the segments a credit analyst would
   draw by hand?
2. Does the proximity structure survive a shift in the sampling window, or does
   the forest reorganise its view of the portfolio from one year to the next?
3. How much of the proximity is explained by the default indicator, and how much
   by protected attributes the model was never given? This is a fairness
   diagnostic that does not require the model to have used the attribute.

## Sketch

```{r}
library(proxima)

rf <- randomForest::randomForest(default ~ ., data = loans, keep.inbag = TRUE)
px <- proximity(rf, newdata = loans, type = "oob")

# Question 1
autoplot(px, type = "network")

# Question 3
permanova(px, ~ default + applicant_race, data = loans)
```
