---
title: "Proximity matrices when n is large"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proximity matrices when n is large}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** The scalability layer is scheduled for phase F3. This vignette
> states the problem and fixes the interface.

A proximity matrix is $n \times n$. At $n = 10^4$ that is 800 MB in double
precision; at $n = 10^5$ it is 80 GB. The ensemble happily fits data at both
sizes, so the matrix, not the model, is what makes the method unusable. This is
not a hypothetical: it is the wall the `e2tree` work hit on the Fannie Mae and
HMDA data.

Three ways round it, in increasing order of how much they give up.

## Sparsity

Most pairs never share a leaf, so most entries are exactly zero. Thresholding
and storing the result as a sparse matrix is lossless for every pair above the
threshold.

```{r}
px <- proximity(rf, newdata = df, sparse = TRUE, threshold = 0.05)
```

## Landmarks and the Nystrom approximation

Compute the proximity exactly on $m \ll n$ landmark observations, then extend it
to the rest by projection:

$$\tilde{P} = P_{n,m}\, P_{m,m}^{-1}\, P_{m,n}.$$

Stratifying the landmark sample on the response keeps the rare class
represented.

```{r}
px_nys <- nystrom(rf, data = df, landmarks = 500, strata = df$y)
```

## Never materialising the matrix

When the quantity of interest is a scalar, a Mantel correlation or a CKA, the
matrix can be traversed in blocks and discarded. The statistic is accumulated in
a streaming fashion and the full $n \times n$ object is never allocated.

```{r}
mantel_test(rf1, rf2, data = df, streaming = TRUE)
```

## Choosing between them

| $n$ | Strategy | Cost |
|---|---|---|
| $\le 5{,}000$ | Dense, exact | Nothing given up |
| $5{,}000$–$50{,}000$ | Sparse, exact above threshold | Small proximities lost |
| $> 50{,}000$ | Nystrom, or streaming statistics | Rank-$m$ approximation |
