---
title: "Proximity matrices when n is large"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proximity matrices when n is large}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
run <- requireNamespace("randomForest", quietly = TRUE)
knitr::opts_chunk$set(eval = run, purl = run)
```

A proximity matrix is $n \times n$. At $n = 10^4$ that is 800 MB in double
precision; at $n = 10^5$ it is 80 GB. The ensemble happily fits data at both
sizes, so the matrix, not the model, is what makes the method unusable. This is
not a hypothetical: it is the wall the `e2tree` work hit on the Fannie Mae and
HMDA data.

Two ways round it are implemented, and they give up different things. A third
is not, and is described at the end for what it is.

```{r setup}
library(Proximum)
library(randomForest)

set.seed(1)
n <- 600
X <- data.frame(matrix(rnorm(n * 6), n, 6))
y <- factor(ifelse(X$X1 + X$X2 + rnorm(n) > 0, "a", "b"))
training <- cbind(X, y = y)

forest <- randomForest(y ~ ., data = training, ntree = 500)
px <- proximity(forest, newdata = training)
format(object.size(px), units = "auto")
```

## Sparsity: keep every value above a threshold

Thresholding and storing the result as a sparse matrix is lossless for every
pair above the threshold and stores nothing for the pairs below it.

```{r sparsify}
sp <- sparsify(px, threshold = 0.05)
sp
```

```{r sparsify-summary}
summary(sp)
```

The saving is real. What it is not is durable:

```{r sparsify-cost}
c(
  dense = format(object.size(px), units = "auto"),
  sparse = format(object.size(sp), units = "auto")
)
```

The result is a `proximity_sparse`, not a `proximity`, and no statistic in the
package will take it:

```{r sparsify-refusal, error = TRUE}
mantel_test(sp, px, n_perm = 99)
```

That refusal is the point rather than an omission. Every statistic here runs on
the induced dissimilarity or on the doubly centred matrix, and both are dense
whatever the proximity was: $1 - P$ turns every stored zero into a one, and the
Gower centring leaves no zero at all.

```{r sparsify-densification}
dissimilarity <- as_dissimilarity(px)
c(
  proximity = mean(as.matrix(px) != 0),
  dissimilarity = mean(dissimilarity != 0),
  centred = mean(double_centre(dissimilarity) != 0)
)
```

So `sparsify()` is a storage format. Use it to hold a matrix between sessions
or to hand it to something outside the package; `as.matrix()` spends the memory
back when you want the statistics.

## Landmarks: the Nystrom approximation

Compute the proximity exactly against $m \ll n$ landmark observations, then
extend it to the rest by projection:

$$\tilde{P} = P_{n,m}\, P_{m,m}^{-1}\, P_{m,n}.$$

Nothing of size $n \times n$ is ever formed, on the way in or on the way out.

```{r nystrom}
approximation <- nystrom(forest, training, landmarks = 60, strata = training$y)
approximation
```

Stratifying the landmark sample on the response is what keeps a rare class
represented, and it is the argument to reach for when one class is small.

```{r nystrom-summary}
summary(approximation)
```

`diagonal error` is the price. The approximation has $\tilde{P}_{ii} \ne 1$,
where an exact proximity has one, and that departure is the cheapest single
measure of how much the landmarks failed to span the sample. It falls as the
landmarks are added:

```{r nystrom-landmarks}
sapply(c(20, 60, 180), function(m) {
  summary(nystrom(forest, training, landmarks = m))$diagonal_error
})
```

### What the approximation is for

Unlike the sparse form, this one survives being used, because the question it
answers is the geometric one. The configuration comes out of the stored factor
at $O(nr^2)$ instead of $O(n^3)$, and `protest()` takes the object directly:

```{r nystrom-embedding}
coordinates <- embedding(approximation, k = 2)
dim(coordinates)

protest(approximation, px, n_perm = 199)
```

The pairwise statistics still refuse it, for the same reason as before: there
is no matrix to correlate entry by entry without building one.

```{r nystrom-refusal, error = TRUE}
cka(approximation, px)
```

## How many trees, and how much does the answer move?

Neither of the above helps if the matrix is unstable, and the number of trees
that settles it is a question with an answer.

```{r n-trees}
required <- n_trees_required(forest, training, eps = 0.2)
required
attr(required, "path")
```

`autoplot()` draws the search it recorded: the criterion against the block
size on log axes, with the target, the answer, and the power law fitted
through the measured points.

```{r n-trees-plot, fig.width = 6, fig.height = 4}
autoplot(required)
```

The answer is an integer and goes on behaving as one, so it can be handed
straight back to the engine that raised the question:

```{r n-trees-integer}
required + 100L
```

The criterion falls as a power of the block size, so a target set too low is
not reached by any ensemble you would fit. When that happens the result is `NA`
and the projection says what it would take:

```{r n-trees-unreachable}
unreachable <- n_trees_required(forest, training, eps = 0.01)
c(answer = unreachable, projected_trees = attr(unreachable, "projected"))
```

`stability()` asks the same question of replicates you already hold, and makes
no assumption about where they came from:

```{r stability}
replicates <- lapply(1:4, function(i) {
  proximity(randomForest(y ~ ., data = training, ntree = 250), newdata = training)
})
agreement <- stability(replicates)
agreement
```

Its plot shows every one of the $R(R-1)/2$ comparisons, with the median and
the percentile interval marked. There is no band around a curve here, because
there is no curve: the object holds one set of dependent agreements, and the
spread of them is the whole of what it can say.

```{r stability-plot, fig.width = 6, fig.height = 3.4}
autoplot(agreement)
```

## Choosing between them

| $n$ | Strategy | Cost |
|---|---|---|
| $\le 5{,}000$ | Dense, exact | Nothing given up |
| $5{,}000$ to $50{,}000$ | Sparse for storage, dense for the statistics | Small proximities lost, memory spent again on use |
| $> 50{,}000$ | Nystrom, for the geometry | Rank-$m$ approximation, no pairwise statistics |

## Not implemented: streaming statistics

When the quantity of interest is a scalar, a Mantel correlation or a CKA, the
matrix could in principle be traversed in blocks and discarded, with the
statistic accumulated as it goes and the full $n \times n$ object never
allocated. `mantel_test()` has no `streaming` argument and this paragraph is
the whole of the feature. It is written down here so that the shape of it is
fixed before anything is built, not to suggest that anything has been.
