---
title: "Proximity matrices as statistical objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proximity matrices as statistical objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_rf <- requireNamespace("randomForest", quietly = TRUE)
```

A random forest partitions the predictor space many times over. Two observations
that keep landing in the same leaf are, as far as the forest is concerned, the
same kind of observation. Collecting that co-occurrence over every tree gives
the *proximity matrix*

$$P_{ij} = \frac{1}{B} \sum_{b=1}^{B} \mathbb{I}\left[\ell_b(x_i) = \ell_b(x_j)\right],$$

where $\ell_b(x)$ is the leaf of tree $b$ reached by $x$.

Most implementations treat $P$ as a by-product, handed back on request and
otherwise ignored. `Proximum` treats it as the object of interest.

## Extracting a proximity matrix

```{r, eval = has_rf}
library(Proximum)

set.seed(1)
rf <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 200, keep.inbag = TRUE
)

px <- proximity(rf, newdata = iris)
px
```

`randomForest` does not keep its training data, so `newdata` has to be supplied.

A warning about the matrix `randomForest` hands back on its own. Its signature
reads `oob.prox = proximity`, so a forest fitted with `proximity = TRUE` and
nothing else stores the **out-of-bag** matrix, not the in-bag one — and the fit
records nothing that says which. `Proximum` recovers the flag from the call and
refuses to relabel the matrix behind your back:

```{r, eval = has_rf, error = TRUE}
rf_stored <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 50, proximity = TRUE
)
proximity(rf_stored) # asks for in-bag; the forest has out-of-bag
```

## Diagnostics

`summary()` reports what the matrix looks like and whether the dissimilarity it
induces can be embedded in a Euclidean space — the condition under which
classical multidimensional scaling of the proximity is exact rather than
approximate.

```{r, eval = has_rf}
summary(px)
```

The matrix is mostly zeros. That is not a defect: most pairs of irises never
share a leaf, and the sparsity is what makes the block structure of the matrix
informative.

## In-bag against out-of-bag

The definition above averages over *all* trees, including the trees that were
fitted on $i$ and $j$. Those trees have seen both observations and are inclined
to separate them correctly, which inflates the proximity of same-class pairs.
Restricting the average to the trees for which both observations are
out-of-bag removes that bias.

```{r, eval = has_rf}
px_oob <- proximity(rf, newdata = iris, type = "oob")
px_oob
```

The two matrices agree closely here, because 200 trees leave every pair
out-of-bag together in plenty of trees:

```{r, eval = has_rf}
cor(px[upper.tri(px)], px_oob[upper.tri(px_oob)])
```

With few trees the out-of-bag denominator can be empty for some pairs. Those
entries come back as `NA`, not as `0`: the forest has no evidence about the
pair, which is a different statement from "the pair is maximally dissimilar".

```{r, eval = has_rf}
set.seed(1)
small <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 3, keep.inbag = TRUE
)
sum(is.na(proximity(small, newdata = iris, type = "oob")))
```

## The out-of-bag proximity is not a kernel

Debiasing the proximity costs something, and the price is worth stating
plainly. Stack the leaf indicators of all $B$ trees into one matrix $Z$, with a
column per leaf. Two observations share a leaf exactly when they agree in that
column, so the in-bag proximity is

$$P = \frac{1}{B} Z Z^{\top},$$

a Gram matrix over $B$. It is therefore positive semi-definite, and
$\sqrt{1 - P}$ is Euclidean. (This identity is also why `Proximum` computes the
matrix with a sparse cross-product instead of a loop over trees: the same
formula that settles the geometry is two orders of magnitude faster to
evaluate.)

The out-of-bag proximity masks $Z$ to the out-of-bag entries and divides by the
number of trees in which each *pair* was jointly out-of-bag. With $M$ the
out-of-bag mask,

$$P^{\text{oob}} = \left( Z_{\text{oob}} Z_{\text{oob}}^{\top} \right) \oslash \left( M M^{\top} \right),$$

an elementwise quotient of two Gram matrices. The Hadamard quotient of two
positive semi-definite matrices need not be positive semi-definite, and here it
is not:

```{r, eval = has_rf}
c(
  inbag = min(eigen(unclass(px), symmetric = TRUE, only.values = TRUE)$values),
  oob   = min(eigen(unclass(px_oob), symmetric = TRUE, only.values = TRUE)$values)
)
```

The in-bag minimum is zero up to rounding; the out-of-bag one is not close to
it. `summary()` reports this rather than letting it pass silently:

```{r, eval = has_rf}
summary(px)$euclidean
summary(px_oob)$euclidean
```

This matters downstream. Classical multidimensional scaling of the out-of-bag
dissimilarity has negative eigenvalues, so its low-dimensional configuration is
an approximation of something that has no exact Euclidean representation, and
any method that assumes a kernel — including centered kernel alignment — needs
an explicit correction first. Choose `type = "oob"` for an unbiased estimate of
the proximities themselves, and `type = "inbag"` when you need the geometry.

When you need both, `make_psd()` projects the out-of-bag matrix onto the cone of
positive semi-definite matrices and records what it did:

```{r, eval = has_rf}
repaired <- make_psd(px_oob, method = "clip")
repaired
summary(repaired)$euclidean
```

Three corrections are offered. `"clip"` discards the negative directions and is
the nearest positive semi-definite matrix in Frobenius norm; `"flip"` keeps them
with their sign reversed; `"shift"` adds a constant to the diagonal, which is
the only one of the three that leaves every off-diagonal proximity exactly where
it was. None is right in general, which is why the choice is yours and is
recorded on the object.

## Other engines

The proximity is a property of the ensemble, not of the package that fitted it.
`Proximum` extracts it from `ranger` with the same definitions and the same
out-of-bag handling:

```{r, eval = requireNamespace("ranger", quietly = TRUE)}
set.seed(1)
rg <- ranger::ranger(Species ~ ., data = iris, num.trees = 200, keep.inbag = TRUE)
proximity(rg, newdata = iris)
```

Because both engines produce the same object, "do two implementations of the
same forest represent the data the same way?" becomes a question the inference
layer can answer, rather than a question nobody can ask.

## From proximity to dissimilarity

```{r, eval = has_rf}
d <- as.dist(px)
mds <- cmdscale(d, k = 2)
plot(mds, col = iris$Species, pch = 19, xlab = "", ylab = "",
     main = "MDS of the forest proximity")
```

## What comes next

The inference layer — comparing two proximity matrices with a Mantel test,
partitioning their variation with PERMANOVA, aligning them with CKA — is the
subject of `vignette("comparing-forests")`. Scaling past a few thousand
observations is the subject of `vignette("large-n")`.
