---
title: "Proximity matrices as statistical objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proximity matrices as statistical objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_rf <- requireNamespace("randomForest", quietly = TRUE)
```

A random forest partitions the predictor space many times over. Two observations
that keep landing in the same leaf are, as far as the forest is concerned, the
same kind of observation. Collecting that co-occurrence over every tree gives
the *proximity matrix*

$$P_{ij} = \frac{1}{B} \sum_{b=1}^{B} \mathbb{I}\left[\ell_b(x_i) = \ell_b(x_j)\right],$$

where $\ell_b(x)$ is the leaf of tree $b$ reached by $x$.

Most implementations treat $P$ as a by-product, handed back on request and
otherwise ignored. `proxima` treats it as the object of interest.

## Extracting a proximity matrix

```{r, eval = has_rf}
library(proxima)

set.seed(1)
rf <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 200, keep.inbag = TRUE
)

px <- proximity(rf, newdata = iris)
px
```

`randomForest` does not keep its training data, so `newdata` has to be supplied.

## Diagnostics

`summary()` reports what the matrix looks like and whether the dissimilarity it
induces can be embedded in a Euclidean space — the condition under which
classical multidimensional scaling of the proximity is exact rather than
approximate.

```{r, eval = has_rf}
summary(px)
```

The matrix is mostly zeros. That is not a defect: most pairs of irises never
share a leaf, and the sparsity is what makes the block structure of the matrix
informative.

## In-bag against out-of-bag

The definition above averages over *all* trees, including the trees that were
fitted on $i$ and $j$. Those trees have seen both observations and are inclined
to separate them correctly, which inflates the proximity of same-class pairs.
Restricting the average to the trees for which both observations are
out-of-bag removes that bias.

```{r, eval = has_rf}
px_oob <- proximity(rf, newdata = iris, type = "oob")
px_oob
```

The two matrices agree closely here, because 200 trees leave every pair
out-of-bag together in plenty of trees:

```{r, eval = has_rf}
cor(px[upper.tri(px)], px_oob[upper.tri(px_oob)])
```

With few trees the out-of-bag denominator can be empty for some pairs. Those
entries come back as `NA`, not as `0`: the forest has no evidence about the
pair, which is a different statement from "the pair is maximally dissimilar".

```{r, eval = has_rf}
set.seed(1)
small <- randomForest::randomForest(
  Species ~ ., data = iris, ntree = 3, keep.inbag = TRUE
)
sum(is.na(proximity(small, newdata = iris, type = "oob")))
```

## The out-of-bag proximity is not a kernel

Debiasing the proximity costs something, and the price is worth stating
plainly. The in-bag proximity is an average of matrices $Z_b Z_b^{\top}$, where
$Z_b$ is the leaf-membership indicator of tree $b$. Each term is a Gram matrix,
so the average is positive semi-definite, and $\sqrt{1 - P}$ is Euclidean.

The out-of-bag proximity divides each entry by a *different* denominator — the
number of trees in which that particular pair was jointly out-of-bag. The result
is no longer an average of Gram matrices and need not be positive
semi-definite:

```{r, eval = has_rf}
c(
  inbag = min(eigen(unclass(px), symmetric = TRUE, only.values = TRUE)$values),
  oob   = min(eigen(unclass(px_oob), symmetric = TRUE, only.values = TRUE)$values)
)
```

The in-bag minimum is zero up to rounding; the out-of-bag one is not close to
it. `summary()` reports this rather than letting it pass silently:

```{r, eval = has_rf}
summary(px)$euclidean
summary(px_oob)$euclidean
```

This matters downstream. Classical multidimensional scaling of the out-of-bag
dissimilarity has negative eigenvalues, so its low-dimensional configuration is
an approximation of something that has no exact Euclidean representation, and
any method that assumes a kernel — including centered kernel alignment — needs
an explicit correction first. Choose `type = "oob"` for an unbiased estimate of
the proximities themselves, and `type = "inbag"` when you need the geometry.

## From proximity to dissimilarity

```{r, eval = has_rf}
d <- as.dist(px)
mds <- cmdscale(d, k = 2)
plot(mds, col = iris$Species, pch = 19, xlab = "", ylab = "",
     main = "MDS of the forest proximity")
```

## What comes next

The inference layer — comparing two proximity matrices with a Mantel test,
partitioning their variation with PERMANOVA, aligning them with CKA — is the
subject of `vignette("comparing-forests")`. Scaling past a few thousand
observations is the subject of `vignette("large-n")`.
